---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message = FALSE,
                      fig.path="figures/",
                      fig.width=8, fig.height=6,
                      fig.show='hold', fig.align='center')

options(width=100)
```

### Bivariate Normal Distribution

We will produce three bivariate normal distribution curves using R.  It would be a good idea to try this program for various values of r between -1 and 1 to explore how the shape of the normal distribution varies with the correlation.

Download the Normal plot R file here: [normplot.R](src/normplot.R)

To simulate multivariate distributions we will create a function that calls other functions from the `MASS` package.  Our function will require an argument for &rho;, `rho`, which will be used to create a bivariate (2 x 2) correlation matrix.  From the correlation matrix two new variables will be simulated using the `mvnorm()` function and their density distributions estimated and plotted with the `kde2d()` and `persp()` functions, respectively.  Our function will also print the correlation coefficient for the two simulated variables.

We will start with two variables with &rho; = 0.

```{r bivariate-density-no-correl}
library(MASS)

biv_dist <- function(rho) {
  rho <- rho
  
  Sigma <- rbind(c(1, rho),
                 c(rho, 1))
  
  set.seed(4960)
  xy <- as.data.frame(mvrnorm(1e5, c(1, 1), Sigma))
  
  den3d <- kde2d(xy$V1, xy$V2, n = 30)
  persp(den3d, box=FALSE)
  
  print(cor(xy$V1, xy$V2))
}

biv_dist(0)
```

The first plot shows the case where the correlation, &rho;, is equal to zero. This special case is called the circular normal distribution. Here, we have a perfectly symmetric bell-shaped curve in three dimensions.

```{r bivariate-density-correl}
biv_dist(0.7)
```

As &rho; increases that bell-shaped curve becomes flattened on the 45-degree line. So for &rho; equals 0.7 we can see that the curve extends out and becomes flattened in the perpendicular direction.

```{r bivariate-density-strong-correl}
biv_dist(0.9)
```

Increasing &rho; to 0.9 the curve becomes broader and the 45-degree line and even flatter still in the perpendicular direction.